---
title: "Validating against the New York Fed ACM series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating against the New York Fed ACM series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, dpi = 120)
options(digits = 4)
```

A term premium estimate is only as trustworthy as the evidence that the code
producing it is correct. The Federal Reserve Bank of New York publishes the
output of the Adrian, Crump and Moench (2013) model — fitted yields, the
risk-neutral component and the term premium, monthly since 1961 — which makes
an unusually strong benchmark available: not a stylised fact to be compared
against loosely, but a specific series to be reproduced.

This vignette reproduces it, then investigates what does not match and why.
Both halves matter. A validation that reports only the agreement is marketing.

```{r setup}
library(termpremia)

panel <- yield_panel(gsw_monthly, units = "percent",
                     maturity_unit = "months",
                     instrument = "government", issuer = "US")

fit <- suppressWarnings(atsm(panel, n_factors = 5))
fit
```

Five factors and a 1–120 month grid, matching the published specification. The
`suppressWarnings()` is deliberate and is explained at the end — the model
reports a real problem with itself.

## The benchmark

`acm_published` holds the New York Fed's own output. Its internal identity is
worth checking first, because it is the identity our implementation must also
satisfy:

```{r}
with(acm_published, max(abs(fitted - (risk_neutral + term_premium))))
```

Exact to machine precision. Ours is too, by construction:

```{r}
max(abs(fit$fitted - (fit$risk_neutral + fit$term_premium)))
```

## Headline comparison

```{r}
mine <- term_premium(fit, maturity = 120)
pub <- acm_published[acm_published$maturity == 120, c("date", "term_premium")]

cmp <- merge(mine, pub, by = "date")
cmp$mine_bp <- cmp$value * 1e4
cmp$pub_bp <- cmp$term_premium * 100

c(observations = nrow(cmp),
  corr_levels  = cor(cmp$mine_bp, cmp$pub_bp),
  corr_changes = cor(diff(cmp$mine_bp), diff(cmp$pub_bp)),
  mean_mine    = mean(cmp$mine_bp),
  mean_published = mean(cmp$pub_bp))
```

Correlation of 0.9997 in levels and 0.9935 in monthly changes, across all 763
month-ends. For scale, Cohen, Hördahl and Xia (2018) report change correlations
of 0.77–0.92 between published implementations of *different* models.

```{r, echo = FALSE}
plot(cmp$date, cmp$pub_bp, type = "l", lwd = 2, col = "grey40",
     xlab = "", ylab = "basis points",
     main = "US 10-year term premium")
lines(cmp$date, cmp$mine_bp, lwd = 1.4, col = "firebrick")
abline(h = 0, col = "grey85", lty = 2)
legend("topright", c("NY Fed published", "termpremia"),
       col = c("grey40", "firebrick"), lwd = c(2, 1.4), bty = "n", cex = 0.85)
```

The model also fits the curve itself closely:

```{r}
sqrt(mean((fit$observed - fit$fitted)^2)) * 1e4  # RMSE, basis points
```

## Where the remaining difference lives

The mean levels differ by about 15bp. Because the New York Fed publishes the
*components*, not just the premium, that gap can be localised rather than
guessed at.

```{r}
idx <- match(120, fit$maturities)
pub10 <- acm_published[acm_published$maturity == 120, ]

data.frame(
  component = c("fitted yield", "risk-neutral", "term premium"),
  mine = c(mean(fit$fitted[, idx]), mean(fit$risk_neutral[, idx]),
           mean(fit$term_premium[, idx])) * 100,
  published = c(mean(pub10$fitted), mean(pub10$risk_neutral),
                mean(pub10$term_premium)),
  mean_abs_diff_bp = c(
    mean(abs(fit$fitted[, idx] * 100 - pub10$fitted)),
    mean(abs(fit$risk_neutral[, idx] * 100 - pub10$risk_neutral)),
    mean(abs(fit$term_premium[, idx] * 100 - pub10$term_premium))
  ) * 100
)
```

**The fitted yields agree to 0.77bp. The entire discrepancy is in the
risk-neutral component.** We do not disagree about fitting the curve; we
disagree about where to cut it.

That localisation is structural, not incidental. Fitted yields depend on the
risk-adjusted combination $(\mu - \lambda_0, \Phi - \lambda_1)$, while
risk-neutral yields depend on $\mu$ and $\Phi$ alone. Agreeing on the former
while differing on the latter means the real-world dynamics differ, with the
prices of risk absorbing the difference.

### Why the real-world dynamics are the fragile half

The two halves of the model are identified by very different amounts of
information. The pricing side is pinned by ~120 yields observed at *every*
date. The real-world dynamics are pinned by the time series alone — and the
factors are close to a unit root:

```{r}
rho <- fit$spectral_radius[["real_world"]]
c(spectral_radius = rho,
  half_life_months = log(0.5) / log(rho),
  observations = length(fit$dates))
```

A shock with a 78-month half-life, observed over 763 months, gives roughly ten
effective independent observations of the mean reversion. That parameter is
barely estimated at all, and the risk-neutral yield is sensitive to it.

## Correcting the small-sample bias

Worse than imprecise, it is biased. Least squares understates autoregressive
persistence in finite samples, and the bias grows as the process approaches a
unit root — exactly the region interest rate factors live in. Bauer, Rudebusch
and Wu (2012) correct it by bootstrap: simulate from a candidate `Phi`,
re-estimate by OLS, and search for the `Phi` whose OLS estimates are centred on
what was actually observed.

```{r}
bc <- suppressWarnings(atsm(panel, n_factors = 5, p_dynamics = "brw"))
bc$brw$spectral_radius[c("ols", "corrected")]
```

The correction is large. In half-life terms:

```{r}
rho2 <- bc$brw$spectral_radius[c("ols", "corrected")]
round(log(0.5) / log(rho2) / 12, 1)   # years
```

Six and a half years becomes effectively permanent. The cross-sectional fit is
untouched — by construction, because the yield curve at a point in time carries
no information about how fast the factors mean-revert, and this implementation
holds the risk-adjusted dynamics fixed and lets the prices of risk absorb the
change:

```{r}
max(abs(bc$fitted - fit$fitted))
```

What moves is the split. A more persistent `Phi` means today's rate level is
expected to last, so the expectations component tracks the level instead of
reverting to a sample average, and the term premium — its residual — becomes
smaller in variance:

```{r}
j <- match(120L, fit$maturities)
rbind(
  ols = c(tp_mean = mean(fit$term_premium[, j]), tp_sd = sd(fit$term_premium[, j]),
          rn_sd = sd(fit$risk_neutral[, j])),
  brw = c(mean(bc$term_premium[, j]), sd(bc$term_premium[, j]),
          sd(bc$risk_neutral[, j]))
) * 1e4
```

That is BRW's substantive claim: a material share of what the uncorrected model
calls term premium variation is an artefact of mean reversion estimated too
fast.

Note what this does to the benchmark comparison:

```{r}
cmp$brw_bp <- bc$term_premium[match(cmp$date, fit$dates), j] * 1e4
c(corr_with_published   = cor(cmp$brw_bp, cmp$pub_bp),
  corr_with_uncorrected = cor(cmp$brw_bp, cmp$mine_bp))
```

```{r, echo = FALSE}
plot(cmp$date, cmp$mine_bp, type = "l", lwd = 1.4, col = "firebrick",
     xlab = "", ylab = "basis points",
     ylim = range(cmp$mine_bp, cmp$brw_bp),
     main = "US 10-year term premium: effect of the bias correction")
lines(cmp$date, cmp$brw_bp, lwd = 1.4, col = "steelblue")
abline(h = 0, col = "grey85", lty = 2)
legend("topright", c("OLS dynamics", "bias-corrected dynamics"),
       col = c("firebrick", "steelblue"), lwd = 1.4, bty = "n", cex = 0.85)
```

The correlation with the New York Fed series falls sharply, and it should: the
published series is uncorrected. Agreement with it is evidence about the
*estimator*, which the previous sections established; it is not evidence about
which specification is right. A bias-corrected model is a different, defensible
answer to the same question, and the two disagree by considerably more than the
15bp level gap investigated above.

The correction is a bootstrap, so it costs a few seconds and is a Monte Carlo
estimate; `brw_control()` fixes a seed by default so that repeated fits on the
same data agree exactly.

## Can the short rate account for it?

One input is a genuine modelling choice rather than a given. The short rate
enters excess returns directly as $-r_t$, so it moves the estimated prices of
risk and therefore the level of the premium.

By default the package uses the panel's own one-month yield. But GSW exclude
Treasury bills and all securities with under three months to maturity, so their
short end is *extrapolated*, and they caution against it. ACM's own stated
inputs include Federal Reserve H.15. The two disagree substantially:

```{r}
fitted_1m <- gsw_monthly$yield[gsw_monthly$maturity == 1]
c(mean_gap_pp = mean(fitted_1m - tbill_3m$value),
  sd_gap_pp = sd(fitted_1m - tbill_3m$value))
```

The fitted short end runs ~28bp above actual bill rates, with a large standard
deviation. Substituting the bill series:

```{r}
fit_bill <- suppressWarnings(
  atsm(panel, n_factors = 5, short_rate = tbill_3m, short_rate_units = "percent")
)

alt <- merge(term_premium(fit_bill, 120), pub, by = "date")

data.frame(
  short_rate = c("GSW fitted 1m", "3m Treasury bill", "NY Fed published"),
  mean_tp_bp = c(mean(cmp$mine_bp), mean(alt$value) * 1e4, mean(cmp$pub_bp)),
  curve_rmse_bp = c(sqrt(mean((fit$observed - fit$fitted)^2)),
                    sqrt(mean((fit_bill$observed - fit_bill$fitted)^2)),
                    NA) * 1e4
)
```

**The two short-rate choices bracket the published series.** The fitted
one-month yield puts the premium 15bp below; an actual bill rate puts it 12bp
above. That is strong evidence the short rate is the lever, and it bounds the
effect at roughly 27bp.

It does not identify the answer. The bill series is a *three*-month rate
standing in for the model's *one*-month period, which is internally
inconsistent and degrades the curve fit by an order of magnitude (from 1.5bp to
14bp RMSE). It is a diagnostic, not a recommended setting. Resolving this
properly needs a genuine one-month rate over the full sample, which the free
sources do not provide before 2001.

## What this implementation does and does not establish

Established:

- The estimator reproduces the published series at 0.9997 in levels and 0.9935
  in changes, and fits the curve to 1.5bp.
- The decomposition identity holds exactly.
- The residual gap lies entirely in the risk-neutral component, and the short
  rate can move it by more than the gap itself.

Not established:

- Which specific choice the New York Fed made. Candidate explanations —
  the short rate, factor scaling, sample conventions — are not separately
  identified by the published output.

Since ACM is fully deterministic, every basis point of difference is
attributable to some concrete difference in inputs or conventions rather than to
sampling noise. That makes the gap findable in principle; it is not findable
from the published series alone.

## A limitation the model reports about itself

The `suppressWarnings()` at the top was hiding this:

```{r}
c(min_expected_short_rate_pct = fit$zlb$min * 100,
  share_negative = fit$zlb$frac_negative)
```

Fitted over a sample spanning the zero lower bound, the model projects expected
short rates to about −3%, a policy rate the United States has never delivered.
Gaussian affine models impose no lower bound, and the BIS review flags exactly
this: near the effective lower bound such models place probability on rates
policy will not deliver, and the resulting term premium is biased.

This is a property of the model class, not a defect in this implementation, and
the package reports it rather than hiding it. Shadow-rate models (Wu and Xia
2016) are the usual remedy and are not implemented here.

## References

Adrian, T., R. K. Crump and E. Moench (2013). "Pricing the term structure with
linear regressions." *Journal of Financial Economics* 110(1), 110–138.

Bauer, M. D., G. D. Rudebusch and J. C. Wu (2012). "Correcting estimation bias
in dynamic term structure models." *Journal of Business & Economic Statistics*
30(3), 454–467.

Cohen, B., P. Hördahl and D. Xia (2018). "Term premia: models and some stylised
facts." *BIS Quarterly Review*, September.

Gürkaynak, R. S., B. Sack and J. H. Wright (2007). "The U.S. Treasury yield
curve: 1961 to the present." *Journal of Monetary Economics* 54(8), 2291–2304.

Wu, J. C. and F. D. Xia (2016). "Measuring the macroeconomic impact of monetary
policy at the zero lower bound." *Journal of Money, Credit and Banking* 48(2–3),
253–291.
